---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 300
)
```

# Tractable Ridge Regression for Paired Comparisons

This document illustrates how to use the R (R Core Team, 2024) code accompanying [Varin and Firth (2024)](https://arxiv.org/abs/2406.09597) to replicate the analysis of the last 28 seasons of English Premier League. The figures in the paper differ in minor respects from those produced in this document due to some manual editing for inclusion in the paper.

<!-- badges: start -->
<!-- badges: end -->

Load some libraries that will be used in this document:
```{r, message = FALSE, warning = FALSE}
library("dplyr")
library("tidyr")
library("ggplot2")
library("signs")
```

Load the code that implements the methodology discussed in [Varin and Firth (2024)](https://arxiv.org/abs/2406.09597):
```{r, warning = FALSE, message = FALSE}
source("peb.R")
```

Ridge regression for paired comparisons with the pairwise empirical Bayes method is implemented in the`peb` function:
```{r, eval = FALSE}
peb <- function(tournament, lambda = NULL, home_effect = FALSE, a = 2 * nlevels(tournament$home), eps = 1e-4, ...)
```
The main argument of `peb` is `tournament`. The code assumes that `tournament` is a `data.frame` with at least three columns called `home`, `away` and `outcome`. Columns `home` and `away` are factors that identify the teams playing at home and away, respectively. Column `outcome` is the outcome of the match coded as `1` in case of victory of the away team, `2` in case of draw and `3` in case of victory of the home team. Further arguments of `peb` are:

- the ridge tuning parameter `lambda`. The default value is `NULL` in which case $\lambda$ is estimated by the pairwise empirical Bayes method;
- the logical variable `home_effect` that indicates whether a home-field effect parameter has to be included (`TRUE`) or not (`FALSE`);
- the small-sample adjustment parameter `a`. The default value for `a` is `2 * levels(tournament$data)` which corresponds in adding one imaginary concordant couple and one imaginary discordant couple of paired comparisons to each of the teams, as discussed in the paper. Estimation without the small-sample adjustment is obtained with `a=0`.

Please refer to [Varin and Firth (2024)](https://arxiv.org/abs/2406.09597) for further details on the meaning of the quantities indicated above.

Load English Premier League results for the seasons 1995-1996 to 2022-2023:
```{r}
load("premier_1995-2023.RData")
```

The loaded object is a `data.frame` with the results of all `r nrow(premier_data)` matches in the 28 seasons from 1995-1996 to 2022-2023:
```{r}
class(premier_data)
head(premier_data)
```
The long-term proportions of home wins, draws and away wins are
```{r}
options(digits = 2)
ref_probs <- prop.table(table(premier_data$outcome))
ref_probs
```

For illustration, let us fit the paired comparison model to season 2015/2016 without home-field effect:
```{r}
season_1516 <- premier_data %>%
  filter(season == "2015-2016") 
fit <- peb(season_1516, home_effect = FALSE)
```
The fitted object is a `list` with various components:
```{r}
class(fit)
names(fit)
```
Among the components of `fit` are the model parameters $\gamma$ and $\delta$:
```{r}
fit$gamma
fit$delta
```
The estimate of $\delta$ is set equal to zero since no home-field effect is assumed.
The tuning parameter $\lambda$ estimated with the pairwise empirical Bayes method is
```{r}
fit$lambda
```
The most relevant component of `fit` is the vector $\mu$ of the estimated strength parameters, below ordered from best to worst rated team:
```{r}
sort(fit$mu, decreasing = TRUE)
```
Now, we refit the model with the inclusion of the home-field effect:
```{r}
fit2 <- peb(season_1516, home_effect = TRUE)
```
The estimates of the model parameters and the tuning parameter become
```{r}
fit2$gamma
fit2$delta
fit2$lambda
```
The corresponding vector of estimated strength parameters is
```{r}
sort(fit2$mu, decreasing = TRUE)
```

Now we load the function that is used to calculate the comparison methods which are maximum likelihood and bias-reduced maximum likelihood. Estimates with these methods are calculated with the `bpolr.R` function of Kosmidis (2014). This function can be downloaded from Ioannis Kosmidis' [web page](https://www.ikosmidis.com/):
```{r}
library("archive")
url_address <- "https://www.ikosmidis.com/files/rssb12025-sup-0002-supplementary-materials-ikosmidis.zip"
archive_extract(url_address, dir = "tmp")
source("tmp/supplementary/bpolr.R")
```

The next lines replicate the forecasting exercises of [Varin and Firth (2024)](https://arxiv.org/abs/2406.09597). For each season, paired comparison models are trained with the first $k$ matchweeks (with $k=10, 15, 20, 25, 30$) and then validated by predicting the rest of the tournament:
```{r, message = FALSE, warning = FALSE}
train_weeks <- seq(10, 30, by = 5)
n_weeks <- length(train_weeks)
methods <- c("PEB", "MLE", "BRMLE")
n_methods <- length(methods)
seasons <- unique(premier_data$season)
n_seasons <- length(seasons)
premier_skill_scores <- data.frame(season = rep(seasons, each = n_methods * n_weeks), week = rep(rep(train_weeks, each = n_methods, n_seasons)), skill_score = NA, method = rep(rep(methods, n_seasons), n_weeks))
premier_skill_scores$method <- factor(premier_skill_scores$method, levels = c("MLE", "BRMLE", "PEB"))

for (s in seasons) {
  for (w in train_weeks) {
    this_season <- premier_data %>% filter(season == s)
    this_season <- droplevels(this_season)
    train_set <- this_season %>% filter(week <= w)
    test_set <- this_season %>% filter(week > w)
    peb_fit <- peb(train_set, home_effect = TRUE, a = 40)
    mle_fit <- brmle(train_set, br = FALSE, maxit = 5000)
    br_fit <- brmle(train_set, br = TRUE, maxit = 5000)
    rows <- which(premier_skill_scores$season == s & 
                  premier_skill_scores$week == w)
    premier_skill_scores[rows, "skill_score"] <- 
      c(compute_skillscore(peb_fit, test_set, "probit", ref_probs),
        compute_skillscore(mle_fit, test_set, "logit", ref_probs),
        compute_skillscore(br_fit, test_set, "logit", ref_probs))
  }
}
```

The results of the forecasting exercises are summarized in Figure 2 of [Varin and Firth (2024)](https://arxiv.org/abs/2406.09597) which reports the boxplot summary of the distributions of the logarithmic skill scores calculated after 10, 15, 20, 25 and 30 weeks of matches:
```{r}
## select a set of colors that are safe for colorblind people:
my_colors <- c("#E69F00", "#D55E00", "#0072B2")
ref_color <- "#999999"

my_theme <- theme_minimal() +
  theme(axis.title.x = element_text(),
        axis.title.y = element_text(angle = 0))
    
boxplot_premier <- function() {
  week_names <- as_labeller(
      c(`10` = "Matchweek 10", 
        `15` = "Matchweek 15",
        `20` = "Matchweek 20", 
        `25` = "Matchweek 25",
        `30` = "Matchweek 30"))

  my_plot <- premier_skill_scores %>% 
    ggplot(aes(y = skill_score, x = method, fill = method)) +
    geom_boxplot(alpha = 0.8) +
    facet_wrap(~week, ncol = 5, labeller = week_names) + 
    geom_hline(aes(yintercept = 0.0), col = ref_color, alpha = 0.8) +
    scale_y_continuous(labels = signs_format(accuracy = .01)) +
    scale_fill_manual(values = my_colors) +
    labs(x = NULL, y = NULL, fill = NULL) + 
    my_theme +
    theme(axis.text.x = element_blank(), 
          legend.position = "bottom")
  ## 
  my_plot
}
boxplot_premier() 
```

## References
 
Kosmidis I. (2014). [Improved estimation in cumulative link models](https://academic.oup.com/jrsssb/article/76/1/169/7075945). *Journal of the Royal Statistical Society Series B: Statistical Methodology* **76**(1), 169--196.

R Core Team (2024). *R: A Language and Environment for Statistical Computing*. R Foundation for Statistical Computing, Vienna, Austria. <https://www.R-project.org/>.


Varin C. and Firth D. (2024). Tractable Ridge Regression for Paired Comparisons. <https://arxiv.org/abs/2406.09597>
